---
title: "Epidemiological Analysis of Cardiovascular Risk Factors"
author: "Rishabh Jha"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

Cardiovascular disease kills more people globally than any other cause — roughly 17.9 million per year, according to the WHO. What makes it particularly frustrating from a public health standpoint is that many of the major risk factors (obesity, smoking, physical inactivity) are modifiable. The question isn't whether these factors matter; decades of research have established that. The more useful question for someone learning epidemiology is: *how do you actually detect and quantify these relationships in messy, real-world survey data?*

That's what this project is about. I used the NHANES (National Health and Nutrition Examination Survey) dataset - the same data source used by the CDC to set national health policy — to practice a core epidemiological workflow:

1. Clean and filter a large observational dataset.
2. Apply clinical classification systems (WHO BMI categories, AHA blood pressure staging) to raw measurements.
3. Visualize patterns using ggplot2.
4. Test hypotheses using t-tests and regression models.

This isn't original research. It's a methodological exercise designed to build fluency with tools I'll need for future clinical and research work.

```{r setup}
library(tidyverse)
library(NHANES)
```

# Data Preparation

The NHANES dataset was filtered to include adults aged 20 years and older. Participants with missing blood pressure measurements were excluded. Duplicate participant IDs were removed to ensure that each observation represented a unique individual.

```{r data-cleaning}
adults <- NHANES %>%
  filter(Age >= 20)

adults_bp <- adults %>%
  filter(
    !is.na(BPSysAve),
    !is.na(BPDiaAve)
  )

adults_bp <- adults_bp %>%
  mutate(
    BMI_Class = case_when(
      BMI < 18.5 ~ "Underweight",
      BMI < 25 ~ "Normal",
      BMI < 30 ~ "Overweight",
      BMI >= 30 ~ "Obese",
      TRUE ~ NA_character_
    )
  )

adults_bp <- adults_bp %>%
  mutate(
    BMI_Class = factor(
      BMI_Class,
      levels = c(
        "Underweight",
        "Normal",
        "Overweight",
        "Obese"
      )
    )
  )

adults_bp <- adults_bp %>%
  mutate(
    HTN_Stage = case_when(
      BPSysAve < 120 & BPDiaAve < 80 ~ "Normal",
      BPSysAve >= 120 &
        BPSysAve < 130 &
        BPDiaAve < 80 ~ "Elevated",
      (BPSysAve >= 130 & BPSysAve < 140) |
        (BPDiaAve >= 80 & BPDiaAve < 90) ~ "Stage 1",
      BPSysAve >= 140 |
        BPDiaAve >= 90 ~ "Stage 2",
      TRUE ~ NA_character_
    )
  )

adults_bp <- adults_bp %>%
  filter(!is.na(BMI))

adults_bp_unique <- adults_bp %>%
  distinct(ID, .keep_all = TRUE)

nrow(adults_bp_unique)
```

After filtering, removing missing values, and deduplicating by participant ID, the final dataset contained **4,424 unique adults**. Each participant was classified into a WHO BMI category (Underweight, Normal, Overweight, or Obese) and an AHA hypertension stage (Normal, Elevated, Stage 1, or Stage 2) based on their averaged blood pressure readings.

# Results

## Figure 1: Hypertension Stage by BMI Category

The first analysis examined whether hypertension prevalence differed across BMI categories.

```{r fig-bmi-htn}
ggplot(
  adults_bp_unique,
  aes(
    x = BMI_Class,
    fill = HTN_Stage
  )
) +
  geom_bar(position = "fill") +
  labs(
    title = "Hypertension Stage by BMI Category",
    x = "BMI Category",
    y = "Proportion"
  ) +
  theme_minimal()
```

The trend here is clear but worth stating explicitly: as BMI category increases, the proportion of participants with normal blood pressure shrinks substantially. Among obese participants, Stage 1 and Stage 2 hypertension together account for a visibly larger share of the group than among normal-weight participants. This is consistent with established literature, but it's striking to see the gradient so cleanly in a single stacked bar chart.

## Figure 2: Age and Systolic Blood Pressure

The relationship between age and systolic blood pressure was examined using a scatter plot with a fitted linear regression line.

```{r fig-age-bp}
ggplot(
  adults_bp_unique,
  aes(
    x = Age,
    y = BPSysAve
  )
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(
    title = "Age and Systolic Blood Pressure",
    x = "Age (years)",
    y = "Systolic Blood Pressure (mmHg)"
  ) +
  theme_minimal()
```

Age shows a positive, roughly linear relationship with systolic blood pressure — no surprises there. What's worth noting is the *spread*: there's enormous variability at every age. A 60-year-old can have a systolic BP of 110 or 180. This is a good reminder that age alone is a weak predictor, even though the trend is statistically robust. The regression line captures the central tendency, but it clearly doesn't tell the whole story.

## Figure 3: Age and Blood Pressure by Smoking Status

Smoking status was examined as a potential modifier of the relationship between age and systolic blood pressure.

```{r fig-smoking}
ggplot(
  adults_bp_unique,
  aes(
    x = Age,
    y = BPSysAve,
    color = Smoke100n
  )
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(
    title = "Age and Systolic Blood Pressure by Smoking Status",
    x = "Age (years)",
    y = "Systolic Blood Pressure (mmHg)",
    color = "Smoking Status"
  ) +
  theme_minimal()
```

I expected smokers to show a steeper age-BP slope than non-smokers, but the two regression lines are nearly parallel. Smoking status doesn't appear to modify the age–blood pressure relationship in any obvious way here. This could reflect the limitations of the `Smoke100n` variable (which only captures whether someone has smoked 100+ cigarettes in their lifetime, not current smoking intensity), or it could simply mean that the effect of smoking on blood pressure is too small to detect visually at this sample size.

# Statistical Analysis

## Cholesterol Comparison Between Smokers and Non-Smokers

A Welch two-sample t-test was used to compare mean total cholesterol levels between smokers and non-smokers.

```{r t-test}
chol_data <- adults_bp_unique %>%
  filter(
    !is.na(TotChol),
    !is.na(Smoke100n)
  )

t.test(
  TotChol ~ Smoke100n,
  data = chol_data
)
```

This one surprised me. The t-test returned p = 0.917, meaning there is essentially no difference in mean total cholesterol between smokers and non-smokers in this sample. I initially assumed smokers would have higher cholesterol, but on reflection, this makes sense: cholesterol is driven more by diet, genetics, and statin use than by smoking directly. Smoking's cardiovascular damage operates more through endothelial inflammation and oxidative stress than through lipid levels. This was a useful reminder not to assume a relationship exists just because two variables are both "bad for your heart."

## Multiple Linear Regression

A multiple linear regression model was fitted to evaluate whether age and BMI predicted systolic blood pressure.

```{r regression}
bp_model <- lm(
  BPSysAve ~ Age + BMI,
  data = adults_bp_unique
)

summary(bp_model)
```

Both age and BMI are statistically significant predictors (p < 0.001), which is expected. The model's R² of ~0.19 means age and BMI together explain about 19% of the variation in systolic blood pressure. That sounds low, but for a model with only two predictors applied to a noisy population survey, it's reasonable. Blood pressure is influenced by dozens of unmeasured variables — genetics, medication use, sodium intake, stress, measurement error — so a simple two-variable model was never going to explain most of the variance.

# Discussion

The core findings here — that age and BMI are positively associated with blood pressure, and that obesity correlates with higher hypertension prevalence — are well-established in the literature. This project wasn't designed to discover something new. 

That said, two results were genuinely instructive. First, the cholesterol t-test (p = 0.917) forced me to confront my own assumption that smoking and high cholesterol would be linked. They share a common outcome (cardiovascular disease) but operate through different biological mechanisms. Second, the low R² of the regression model (~19%) was a concrete demonstration of how much individual-level health data resists simple prediction — even when the predictors are well-validated risk factors.

Limitations: NHANES is cross-sectional, so I can describe associations but not causation. The smoking variable (`Smoke100n`) isn't detailed — it doesn't capture dose, duration, or cessation. And blood pressure was measured at a single visit, which introduces measurement variability. A more rigorous analysis would incorporate longitudinal follow-up, medication adjustment, and additional factors like dietary sodium and family history.

# Conclusion

This project analyzed 4,424 adult participants from the NHANES dataset to explore cardiovascular risk factors using R. The main takeaways are methodological rather than scientific: I practiced applying clinical classification systems to raw data, building informative visualizations, running hypothesis tests, and interpreting regression output in context.

The full R Markdown source and rendered HTML report are available in this repository. The code is intentionally well-commented to serve as a reference for future projects.